Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-10-28 paper.tex
as of 10-25
obs. study & mediation sections

\documentclass{article}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

@@

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)

@

<<runStuff,include=FALSE,eval=FALSE>>=
source('R/dataAnalysis.r')
big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
save(list=ls()[!big],file='output/analysis.RData')
@

<<loadStuff,include=FALSE>>=
load('output/analysis.RData')
@

\begin{document}
\maketitle

@@

Over the course of the effectiveness study, Carnegie Learning gathered
log data for each worked problem.
The resulting dataset lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
Treatment schools with mastery data missing
for 90\% or more students were omitted from the analysis, along with
their entire randomization block.
Of the remaining 2,390 students, 87\% (2,091) had log
data.
The students with missing log data were dropped from the observational
study and mediation analysis, but included in the principal
stratification study.
Problem log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

All told, the main analysis included $n=$5308 students, 2390 of whom were assigned to the CTA1 condition and 2918 of whom were assigned to control.
The students were nested within 116 teachers, in 43 schools across five states.

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer, instead of on paper, as is typical.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

There are good reasons to believe that the availability of
hints---help on demand---may increase student learning, but there is
also reason to be skeptical \cite[See][for an overview of the theory]{aleven2016help}.
First of all, as, e.g., \citet{anderson1995cognitive} points out, ``if
students never receive help of any sort, they are in danger of becoming
permanently stuck on some problem'' (p. 190).
That is, hints give students who are stuck a way to continue working
after getting stuck.
Hints may also boost learning by ``helping students identify relevant features in
problems'' \citep[][p. 6]{aleven2016help},  that is, point students
towards aspects of the problem that are ``most important for the goals
of learning'' \citep[][p. 782]{koedinger2012knowledge}.
The final available ``bottom out'' hint typically gives the student
the answer.
Students who bottom out then see a worked example, which can be an
effective tool for learning \citep[e.g.][]{sweller1985use}.
Finally, hints can help improve students' meta-cognitive states, such
as their self-knowledge of skills.

On the other hand, all of these theoretical mechanisms rely on
students to play their part.
In order for hints to be instructive, students need to reflect on
their content, for instance with ``self-explanation''
\citep[c.f.][]{renkl1998learning}.
In the same vein, hints need to be well-designed in order to be
effective \citep{mckendree1990effective}.
This suggests that hints may be helpful for some students but not for
others.
They may even be harmful: as \citet{koedinger2007exploring} put it
(p. 241),
``many lines of research and theory suggest the importance of \dots
withholding information from students so that they can exercise, test,
or reason toward new knowledge on their own.''
\citet{paas1994variability} interpret this dilemma in terms of balancing
``cognitive load,'' that is, optimally allocating a student's
attention to the most relevant aspects of a problem.

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or third ``bottom out''  hint available for a given problem).
They found that, indeed, the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the bottom out hint.

\subsection{Measuring Hint Usage in the CTA1 Data}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package.
Like \citet{beck2008does} and \citet{goldin2012learner}, but unlike
\citet{anderson1995cognitive} and \citet{singh2011feedback}, hint
usage was not randomized.
On the other hand, the CTA1 study includes a posttest, like the
randomized studies but unlike the observational studies cited here.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a post-test.

The CTA1 study was of a much longer duration than
any study of hint effects that we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.
What is relevant for our purposes is that we may narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.

Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
(More mundanely, we do not have access to data
describing the number of hints available for every problem, which
would hinder interpretation of any other measure of hint usage.)

Lastly, any observational study of hint usage must contend with the
fact that students are unlikely to request hints on problems that they
already know how to solve (other than by accident).
Hence, any measure of hint usage is necessarily also a measure of
student ability---students who know more algebra will almost certainly
request fewer hints.
This leads to two related problems: first, student ability
confounds any statistical relationship between hint request frequency
and posttest scores.
Secondly,
a student's decision to request a hint implies both that she
found the problem at least somewhat difficult, and that, in this case,
she responded to that difficulty by requesting a hint.
Our interest is in the second implication, not the first.
Conversely, the significance of a student's decision to forgo a hint
depends on how challenging he or she finds the problem.
Hint request frequency combines students' hint requests (or non-requests) from both
problems they find challenging and trivial, so is a poor measure of
students' use of the hint feature.

@@

This measure is inversely related to the ratio of the percentage (of all
problems) a student makes an error without having requested a hint,
to the percentage of problems a student requests a hint
\footnote{Technically, if $h$ is the event that a student requests
a hint on a problem and $e$ is the event that the student makes
an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(h\text{ and not }e)/Pr(h)\}$}.
Making an error on a problem without having requested a hint reflects
a hesitance to request a hint when one might have been warranted.
$\bar{h}$ ignores the class of challenging problems on which students
are challenged and figure the answer out (or guess correctly) without
requesting a hint.
This class of problems is inherently interesting to our question;
however, it is unidentified.

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)} of students as
high-hint users, so we chose $c=0.6$, approximately the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

Let $\bm{x}_i$ represent a vector of baseline covariates for subject $i$.
In our study of hints, $\bm{x}$ includes indicators for state, school
id, class id, grade (9th or higher), race (White/Asian,
Black/Multi-Racial, or Hispanic/American Indian/Alaskan Native),
special education status, economic disadvantage, and English language
learner status, as well as pre-test scores.
Then we assume \citep[c.f.][]{rosenbaum1983central}:
\begin{ass}{Strong Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |\bm{x}
\end{equation*}
\end{ass}
that conditional on $\bm{x}$, potential outcomes $Y(H=1)$ and $Y(H=0)$
are independent of realized $H$.
Under strong ignorability, one may compare subjects with identical
covariates $\bm{x}$ to estimate causal effects.

Unfortunately, this sort of exact matching is impossible in our finite
sample.
Instead, we estimated ``propensity scores'' \citep{rosenbaum1983central}:
\begin{equation*}
\pi_i=Pr(H_i=1|\bm{x})
\end{equation*}
the probability of a treatment-group subject requesting frequent hints, conditional on
his or her covariate vector $\bm{x}$.
\citet{rosenbaum1983central} shows that conditioning on $\pi$ is
equivalent to conditioning on $\bm{x}$.
We estimated $\pi$ using a Bayesian multilevel logistic regression,
with weakly informative priors, using the \texttt{rstanarm} package in
\texttt{R} \citep{rstanarm}.
The propensity score model was:
\begin{equation*}
\pi_i=logit^{-1}(\alpha+\bm{\tilde{x}}\bm{\beta}+\epsilon^{state}+\epsilon^{school}+\epsilon^{class})
\end{equation*}
where $\bm{\tilde{x}}$ is the $\bm{x}$ vector excluding indicators for
state, school, and class, which were each included using random
intercepts modeled as normally distributed.

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 30 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 21 $H=0$ subjects to a
single $H=1$ subject.

Table \ref{fig:balance} in the online appendix shows that
matching largely eliminated mean differences in pre-treatment
covariates between $H=1$ and $H=0$ students.
The only notable exception is in pretest scores: $H=1$ students tended
to have slightly higher pretest scores than their matched comparisons
with $H=0$.
The p-value from an omnibus covariate balance test
\citep{covBal} is 0.26.

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with linear and quadratic terms for pre-test (the standard error
and confidence interval use the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

\begin{table}
\centering
\begin{tabular}{crllcll}
&&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
\hline
\multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
&ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
&TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
&ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
\hline
\multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
&[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
\hline
\end{tabular}
\caption{Estimates with and without regression adjustment and
  sensitivity analysis of the effect of hint usage.}
\label{tab:matchResults}
\end{table}

@@

We found that hint requests have a negative effect: students who tend
to request hints more often than the median score lower on the
posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after control for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

>>>>

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle

@@

Over the course of the effectiveness study, Carnegie Learning gathered
log data for each worked problem.
The resulting dataset lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
Treatment schools with mastery data missing
for 90\% or more students were omitted from the analysis, along with
their entire randomization block.
Of the remaining 2390 students, 87\% (2,091) had log
data.
For the sake of simplicity, the treatment students without log data
were dropped from the study.\footnote{Including these students in a
  principal stratification model is straightforward \citep{aoas}; see
  the online supplement for results from a model including them, which
  were broadly similar to the results presented here. Including
  subjects with missing log data in a mediation or observational study
  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
Problem log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
The students were nested within 116 teachers, in 43 schools across five states.

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

There are good reasons to believe that the availability of
hints---help on demand---may increase student learning, but there is
also reason to be skeptical \cite[See][for an overview of the theory]{aleven2016help}.
First of all, as, e.g., \citet{anderson1995cognitive} points out, ``if
students never receive help of any sort, they are in danger of becoming
permanently stuck on some problem'' (p. 190).
That is, hints give students who are stuck a way to continue working.
Hints may also boost learning by ``helping students identify relevant features in
problems'' \citep[][p. 6]{aleven2016help},  that is, point students
towards aspects of the problem that are ``most important for the goals
of learning'' \citep[][p. 782]{koedinger2012knowledge}.
Hints can help improve students' self-awareness and
problem-solving strategy, such as by clarifying which skills they have
and have not mastered.

Most problems are associated with several hints, arranged in a
sequence so that a student who remains stuck after one hint may
request another.
The last hint in the sequence will show a complete solution to the
problem---essentially showing students a worked example, which can be an
effective tool for learning \citep[e.g.][]{sweller1985use}.

On the other hand, all of these theoretical mechanisms rely on
students to play their part.
In order for hints to be instructive, students need to reflect on
their content in productive ways.
This suggests that hints may be helpful for some students but not for
others.
In the same vein, hints need to be well-designed in order to be
effective \citep{mckendree1990effective}, so some hints may be more
helpful than others.

Hints may even be harmful: as \citet{koedinger2007exploring} put it
(p. 241),
``many lines of research and theory suggest the importance of \dots
withholding information from students so that they can exercise, test,
or reason toward new knowledge on their own.''
\citet{paas1994variability} interpret this dilemma in terms of balancing
``cognitive load,'' that is, optimally allocating a student's
attention to the most relevant aspects of a problem.

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or final hint available for a given problem).
They found that, indeed, the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the final hint.

\subsection{Measuring Hint Usage in the CTA1 Data}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package, including the associated textbook and
recommended classroom practices.
Access to the full CTA1 curriculum was randomized, rather than any
specific aspect of usage, and effectiveness was measured with a
standardized posttest covering a broad range of Algebra I skills.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a posttest.

The CTA1 study was of a much longer duration than
any of the studies of hint effects we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.
What is relevant for our purposes is that we may narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.

Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
% (More mundanely, we do not have access to data
% describing the number of hints available for every problem, which
% would hinder interpretation of any other measure of hint usage.)

Lastly, any observational study of hint usage must contend with the
fact that students are unlikely to request hints on problems that they
already know how to solve.
Hence, any measure of hint usage is necessarily also a measure of
student ability---students who know more algebra will almost certainly
request fewer hints.
This leads to two related problems: first, student ability
confounds any statistical relationship between hint request frequency
and posttest scores.
Secondly,
a student's decision to request a hint implies both that she
found the problem at least somewhat difficult, and that, in this case,
she responded to that difficulty by requesting a hint.
Our interest is in the second implication, not the first.
Conversely, the significance of a student's decision to forgo a hint
depends in part on how challenging he or she finds the problem.
Hint request frequency combines students' hint requests (or non-requests) from both
problems they find challenging and trivial, so is a poor measure of
students' disposition to use the hint feature.

@@

This measure is inversely related to the ratio of the percentage (of all
problems) a student makes an error without having requested a hint,
to the percentage of problems a student requests a hint.%
\footnote{Technically, if $h$ is the event that a student requests
a hint on a problem and $e$ is the event that the student makes
an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(e\text{ and not }h)/Pr(h)\}$.}
Making an error on a problem without having requested a hint reflects
a hesitance to request a hint when one might have been warranted.
$\bar{h}$ ignores the class of challenging problems on which students
are challenged and figure the answer out (or guess correctly) without
requesting a hint.
This class of problems is inherently interesting to our question;
however, it is unidentified.

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately 30\% of students as
high-hint users, so we chose $c$ as the
70th percentile of $\bar{h}$,
approximately 0.6.
This definition agreed with the model's classification in
approximately 93\% of cases.

@@

Let $\bm{x}_i$ represent a vector of baseline covariates for subject $i$.
In our study of hints, $\bm{x}$ includes indicators for state, school
id, class id, and the variables described in Table \ref{tab:covariateBalance}.
Then we assume \citep[c.f.][]{rosenbaum1983central}:
\begin{ass}{Strong Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |\bm{x}
\end{equation*}
\end{ass}
that conditional on $\bm{x}$, potential outcomes $Y(H=1)$ and $Y(H=0)$
are independent of realized $H$.
Under strong ignorability, one may compare subjects with identical
covariates $\bm{x}$ to estimate causal effects.

Unfortunately, this sort of exact matching is impossible in our finite
sample.
Instead, we estimated ``propensity scores'' \citep{rosenbaum1983central}:
\begin{equation*}
\pi_i=Pr(H_i=1|\bm{x})
\end{equation*}
the probability of a treatment-group subject requesting frequent hints, conditional on
his or her covariate vector $\bm{x}$.
\citet{rosenbaum1983central} shows that conditioning on $\pi$ is
equivalent to conditioning on $\bm{x}$.
We estimated $\pi$ using a Bayesian multilevel logistic regression,
using the \texttt{lme4} package in
\texttt{R} \citep{lme4}.
The propensity score model was:
\begin{equation*}
\pi_i=logit^{-1}(\alpha+\bm{\tilde{x}}\bm{\beta}+\epsilon^{state}+\epsilon^{school}+\epsilon^{class})
\end{equation*}
where $\bm{\tilde{x}}$ is the $\bm{x}$ vector excluding indicators for
state, school, and class, which were each included using random
intercepts modeled as normally distributed, including missingness
indicators for grade, race, sex, and economic disadvantage, and
a natural spline with five degrees of freedom for pretest.

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 44 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 117 $H=0$ subjects to a
single $H=1$ subject.

Figure \ref{fig:balance} and Table \ref{tab:balance} in the online appendix show that
matching largely eliminated mean differences in pre-treatment
covariates between $H=1$ and $H=0$ students.
The only notable exception is in pretest scores: $H=1$ students tended
to have slightly higher pretest scores than their matched comparisons
with $H=0$.
The p-value from an omnibus covariate balance test
\citep{covBal} is 0.39.

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest (the standard error
and confidence interval using the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Oct 28 14:48:44 2019
\begin{table}[ht]
\centering
\begin{tabular}{llrrlrr}
  \hline
  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
  \hline
\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
  \hline &  &  &  &  &  &  \\ 
  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
   \hline
\end{tabular}
\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

% \begin{table}
% \centering
% \begin{tabular}{crllcll}
% &&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
% \hline
% \multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
% &ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
% &TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
% &ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
% \hline
% \multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
% &[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
% \hline
% \end{tabular}
% \caption{Estimates with and without regression adjustment and
%   sensitivity analysis of the effect of hint usage.}
% \label{tab:matchResults}
% \end{table}

@@

We found that hint requests have a negative effect: students who tend
to request hints more often than the median score lower on the
posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after controlling for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.
